---
title: "Community Detection in the Setting of Generalized Random Dot Product Graphs"
author: "John Koo"
geometry: [top=1in, bottom=1in, right=1in, left=1.5in]
day: "1"
month: "December"
year: "2022"
degree: "Doctor of Philosophy"
field: "Statistical Science"
department: "Statistics"
uni: "Indiana University"
location: "Bloomington, IN"
chair: "Michael W. Trosset, Ph.D."
committee2: "Minh Tang, Ph.D."
committee3: "Julia Fukuyama, Ph.D."
committee4: "Roni Khardon, Ph.D."
committee5: "Fangzheng Xie, Ph.D."
bibliography: yourbibfile.bib
multipaper: true
output:
  dissertateUSU::dissertateUSU:
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args: [ "--csl", "apa7.csl" ]
---

```{r global_options, include=FALSE}
## This saves you from having to do this for every chunk
knitr::opts_chunk$set(fig.path = 'figures/',
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
library(dissertateUSU)
```


<!-- Abstract -->
\newpage
\pagestyle{plain} <!-- can use \pagestyle{fancy} for a horizontal line in header -->
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\chapter*{ABSTRACT}
\addcontentsline{toc}{section}{Abstract}

<!-- \abstracttitle -->
\doublespacing

Your abstract words go here. The line below puts the total number of pages at the end of your abstract (as required).

\begin{flushright}(\pageref{LastPage} pages)\end{flushright}


<!-- Dedication -->
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\chapter*{DEDICATION}
\addcontentsline{toc}{section}{Dedication}

Dedicate it.


<!-- Acknowledgments -->
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\chapter*{ACKNOWLEDGMENTS}
\addcontentsline{toc}{section}{Acknowledgments}

Acknowledge those acknowledgable individuals and things.


<!-- Front Matter -->
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\tableofcontents


\newpage
\pagenumbering{arabic}

<!-- Chapter 1 -->
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter{Block Models for Community Detection}

```{r, child="Chapter1.Rmd"}
```


<!-- Chapter 2 -->
\FloatBarrier
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter{Random Dot Product Graphs and Generalized Random Dot Product Graphs}

```{r, child="Chapter2.Rmd"}
```


<!-- Chapter 3 -->
\FloatBarrier
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter{Popularity Adjusted Block Models are Generalized Random Dot Product Graphs}

```{r, child="Chapter3.Rmd"}
```


<!-- Chapter 4 -->
\FloatBarrier
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter{Generalized Random Dot Product Graphs with Community Structure}

```{r, child="Chapter4.Rmd"}
```


<!-- References -->
\FloatBarrier
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter*{REFERENCES}

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.4in}
\setlength{\parskip}{6pt}
\noindent
